---
title: A Travel Mode Detection approach based on hierarchical and non hierarchical clustering
subtitle: Patterns and Trends in Environmental Data
author: Lucia Scheele and Laura Vetter
format:
  html:
    code-fold: true # makes the code in the output collapsable
execute:
  warning: false # hides warnings from the generated output
  message: false        # hides messages from the generated output
lang: en  # switch to "de" if you write your report in german
bibliography: bibliography.bib
---

```{r}
#| code-summary: preprocessing #name des Code Chunks

#loading all the data from the preprocessing files
load("my_environment.rdata")
#load("selected_tracks.rda") (besser weil man dann nur lädt was man wirklich braucht)

library(tmap) #to show data in map viewer, is necesary in 

``` 

## Abstract

This project aims to analyze travel behaviour and clustering of travel segments by different clustering methods and visual analysis. 
Parameters like speed, acceleration, elevation and sinuosity and derived parameters were used for clustering with kmeans and hmeans. Simultaneously, a segmentation by threshold (using the parameter speed) and a manual segmentation using GIS were carried out to compare them to the clustering. 
Verification took place using Chi square test and ANOVA ?????

The results show that 

- 5 clusters, which corresponds to the actual amount of travel modes (+standing)
- hmeans predicts better than kmeans 
- the ward method in hmeans is better than ...
- xxx showed a better (or equally good) result than manual classification. 
-  Verification method 
- ...
BEARBEITEN 

## Introduction
BEARBEITEN 

In order to understand travel behaviour, it is critical to analyze movement data and be able to classify different types of movement correctly (@sadeghian2022stepwise). 
However, the exact detection of the corresponding travel modes is complex and can be time consuming.  
Generic methodologies to detect travel modes have been described using unsupervised learning algorithms, GIS multi-criteria process (eg. by @sadeghian2022stepwise), and supervised learning algorithms or decision trees and confusion matrix(@shamoun2012sensor).
This project aims to answer the reasearch question of which travel modes can be differentiated in trajectories with different travel modes within the trajectories using hierarchical and non-hierarical clustering?
Therefore, the project focuses on a comparison of these two clustering methods, and a comparison with a manual segmentation using ArcGIS and a threshold segmentation using the parameter speed.


Auch reinbringen?:
@laube2011 investigate how temporal scale affects the calculation of movement parameters (speed, sinuosity and turning angle) of animal trajectories. 



## Material and Methods

### Data
The data sample consists of movement data collected with the Tracking App Strava, by two individuals over a timespan of April - May 2024.
Overall, 31 trips of different travel modes were recorded with the movement types walking, running, cycling, bus, tram, train and car. For simplification of the  design and to be able to focus on the application of the clustering methods, a sample of 2 trips with different movement types within the trajectory were selected to be analyzed.
Data points were recorded every second, resulting in a relatively high resolution of data. However, data frequency is not consistent in all cases due to lack of signal (e.g. in a train).
At first, including contextual data such as traffic infrastructure data by the city of Zurich ( [https://www.stadt-zuerich.ch/geodaten/](#0){.uri}) was considered. Out of two reasons, it was decided not to include this data: 1) categorical data is not suitable for the evaluation with the k-means clustering algorithm and 2) the infrastructure data dit not overlap with the trajectories in many cases and was therefore not especially meaningful. Therefore, the focus was set on the analysis of the numerical data.


### Classification of travel modes   

The analysis was made with data of the individual fixes. instead of using segments for the analysis of e.g. acceleration, the data of moving windows ( X steps before and after the fix) was assigned to the individual fix. This was done to have more information for each fix (data point). 

5 different travel modes could be identified in the samples (standing, walking, running, tram, biking), therefore the optimal number of clusters expected is 5, which is also the number of classes assigned in GIS and with the speed parameter. Also, the trajectory data will be analysed with partitioning methods to determine the number of clusters needed, and results will be compared.
Space and time of trajectories were treated to be relative and are filtered out for the analysis. 
The parameters for clustering will be calculated from the given geographical data and include:
- distance point1 - point2
- speed p1 - p2
- acceleration p1 - p2
- mean speed in moving window 
- max speed in moving window
- mean acceleration in moving window 
- max acceleration in moving window
- elevation change in moving window 
- sinuosity in moving window

The calculation of sinuosity, some adaptions needed to be made as some values had the value "Inf". This occurs when the direct distance = 0, meaning that start and end point of the moving window are the same. The actual moving distance is >1, as within the moving window movement occured. 

actual moving distance /direct distance = x/0=Inf  

There are several ways to deal with this problem: 
1) Set inf values to the maximum value of the sinuosity 
2) Filter out values completely from the table
3) Insert the actual distance traveled per 10 points,  actual/direct=x/x=1 

The third method was chosen, resulting in the lowest sinuosity value possible (=1).

#### Classification via threshold  

In order to verify how well speed alone can already predict the different movement types, a classification was made based on speed as the only criterion. The different travel modes were assigned to different speeds where 
 speed_kmh == 0 ~ "1", #standing
    speed_kmh < 5 ~ "2", #walking
    speed_kmh >= 5 & speed_kmh < 18 ~ "3",  #running
    speed_kmh >= 18 & speed_kmh < 30 ~ "4", # velo 
    speed_kmh > 30 ~ "5" # tram

The encountered problem was the unhomogenity of some fixes, e.g. when the threshold is set at a point where the speed varies constantly (e.g. speed>5 = running, speed<5 = walking)
This issue was resolved with manual classification in GIS.


### Verification with manual classification in GIS

In order to test the results of our cluster analysis, it was aimed to assess in how far the cluster categories are consistent with the actual movement types conducted during the trajectory recordings. This required two independent enrichments of the fix point data. In one step, the information of the actual movement type had to be added to the single trajectory fixes by using the recorder's knowledge of the track. This step was conducted within ArcGis (Tools: Select and Calculate Field). To help the recognition of movement type change, the fix points were colored using the speed (km/h) variable.  
In another step, the individual cluster groups were inspected visually and statistically to find out which movement they are likely to represent.

#### kmeans
K means is one of the most commonly used partition based algorithms (@yuan2017review), where k defines the number of clusters and is identified before processing. Each partition is one cluster and contains at least one data point, where each data point is assigned to only one cluster. 
First, partitioning (determining the appropriate  number of k) is carried out using the elbow method from the package factoextra ([https://www.datanovia.com/en/lessons/determining-the-optimal-number-of-clusters-3-must-know-methods/#elbow-method](#0){.uri}) and the cascade method from the package vegan.  
All parameters included in the clustering analysis are numerical:
```{r}
head(km_no_geom, 5)
``` 


#### hmeans
(LUCIA?)

chaining erklären (deshlab ward am besten)


## Results
FEHLER: WARUM ERSCHEINT ein 3ter Track im Label (aber dann nicht auf der Map)?

2 trajectories were chosen which include several travel modes. Trip 1 is a mixture of walking, running, tram and standing (at the tram stop). Trip 2 consists of biking, walking and standing. 
```{r}
tmap_mode("view")
tm_shape(selected_tracks_na_omit)+ 
  tm_dots(col = "trajID", palette = "Paired") +  
  tm_view(set.view = c(8.520515, 47.388322,  16))

```

### Speed and Speed Classification

```{r}
tmap_arrange(speed_map, cluster_speed_map )

```
Here we can see the 
### Speed parameter Classification and GIS Classification

```{r}
#| fig-cap: "Map of the manually assigned groups and the Speed parameter Classification"
#| label: GIS_Group and manual

  tmap_arrange(cluster_speed_map, cluster_GIS_map )
```
Looking at the two 

### k means  (non hierarchical clustering)

#### Partitioning 
In order to define the appropriate number of clusters, the elbow statistics is used. The  bend (like a bent elbow) in the line indicates  the number of appropriate k. 
```{r}
plot_k_elbow

```
This method indicates that the best amount of clusters seems to be 2, as the curve starts to bend there. The method was tried with several configurations (different n-starts, leaving out parameters like elevation or sinuosity, taking one track only), but the indicated number for k stayed the same.
As this does not correspond to the actual amount of clusters we want to find, we try a different method, the cascade (SOURCE LUCIA). Here, we examine the suitability from 2-8 cluster groups.
The SSE (sum of squared errors) is a measure of how well the data points are summarized within the clusters. The lower, the better the clustering. In the ssi (silhouette index) the higher the value (from -1 to 1) the better the clustering.
```{r}
cascade_results
```
The ssi is highest for 5 groups, while the SSE decreases constantly with the rising number of groups. This indicates, that a well chosen amount of k is 5. 

#### applying k means

The importance of the nstart criterium in the kmeans analysis is not to underestimate. In the following graphic, we can see the different results when using no predetermined n, n=20 and n=100.
```{r}
library("gridExtra")
grid.arrange(plot_cluster_5, plot_cluster_5_20, plot_cluster_5_100, ncol = 3)

```
Here we can see the classification of the sample points in the point cloud, how many data points belong to which cluster. From this, it is not yet possible to determine which cluster might correspond to which travel mode and which n suits our classification best. 

```{r}
tmap_arrange( cluster_k5_map, cluster_k5_100_map )
```
Looking at the two clusters k=5 and k5 n=100, we can see that the distribution of clusters is very similar, different to what the plots before indicated. 
However, the clustering does not seem to work perfectly.

```{r}
cluster_k4_map
```
Changing to k=4, running, walking and a third category can be relatively clearly distinguished, but there is no distinction between tram and bike visible. Also curves with the bike (which imply slowing down) or tram stops are not classified correctly. 

###  hmeans (hierarchical  clustering)(LUCIA)
meaning that e.g. cluster bike is closer to running and tram is closer to train 

```{r}
#ward_plot_5
#ward_dendro 
```

```{r}
#P_ward_4

#tmap_arrange(cluster_h5_ward,cluster_k5_map )

```
Works well with differentiation between running and walking, but nbike and tram look the same. 

### Verification

Premises:
Clusterings with different k values are not directly comparable, as they have fundamentally different structures. Therefore, we will focus on the comparison of k=5 only. 
We will compare:
GIS - Speed
GIS - kmeans
GIS - kmeans_100
GIS - hmeans_ward 
speed - k_means_100 ?
speed - hmeans_ward

As we have several categorical variables (cluster groups) and do not suppose a causality between them, we will use  the chi square or fishers test. 


Kontrollgruppe (GIS)  und Clustergruppen 
H0: Die Gruppenzuordnungen (Cluster) sind gleich bei Kontroll und Clustergruppen
H1: Die Gruppenzuordnungen (Cluster) sind ungleich bei Kontroll und Clustergruppen


#### chi square test
```{r}
#chi_man_k4
```
Hier kommt eine sehr niedrige Signifikanz raus, das heisst, die beiden clustermethoden sind sehr unterschiedlich. Warum? Weil sie nicht die gleichen Werte für die gleichen Cluster nehmen. Es ist also notwendig, erst die cluster visuell zu überprüfen abzugleichen und dann zuzuweisen, und dann den Test zu wiederholen. 
Oder wir probieren es mit Korrelation



#### t test mit rID Segmentierung (WAS IST DAS NOCHMAL? )

Erwartetes Ergebnis: der p Wert ist höher, d.h. unsignifikanter als bei t-test ohne rID segmentierung, das heisst die Gruppenzuweisungen sind sich mit rID segmentierung ähnlicher.



## Discussion

It was observed that clustering with single tracks works better than clustering with all tracks. Therefor, the applicability of this method is questionable for bigger datasets or datasets with different movement patterns. 
Furthermore, it needs to be mentioned that for thevalidation with GIS, the  speed information was used as a control. Therefore, independent control of cluster quality  cannot be guaranteed. 




Left for further analysis:
- At first, the turning angle was considered as an alternative parameter to mitigate the inaccuracies that occured when the bike slowed down in curves, however there was no time left to include this in the analysis and will therefore be left for another project. 
- another interesting investigation would have been the possibility to extract the parameters by which the clusters were formed (how big of a weight had, e.g. the speed in comparison to the average acceleration). If this were possible, the weights could be adjusted to balance certain inaccuracies as well. 
-  Further clustering, such as kmodes, which also allows ordinal/categorical variables to be considered, would have been equally interesting and will be thought about in the future. 



UNFINISHED, to be included in discussion:
- The same goes for the spatial proximity of the data. This might have led to a disadvantage for clustering? (Lucia)
- scale/normalisation. Warum normalisieren wir die Werte? Gibt es Nachteile bei einer Normalisierung(look up dengler?)?
- was haben wir nochmal mit den Randpunkten gemacht? Also wenns von einem track zum nächsten geht? haben wor dafür eine Function oder so damit die speeds etc nicht verfälscht sind?
- first clustering with only one track & verification with CHi/ANOVA and then with various & verification. -> this tells us wether method is generalizable or not? 


## Appendix

### Wordcount

<!-- after installing the wordcountaddin, remove the line "#| eval: false" -->

```{r}
#for word count:
#install.packages("pacman") 
library("pacman")
p_install_gh("benmarwick/wordcountaddin")
wordcountaddin::word_count("index.qmd") 

#knitr::spin("preprocessing.R") --> macht "run" von preprocessing file und produziert html. 

```





