---
title: A Travel Mode Detection approach based on hierarchical and non hierarchical clustering
subtitle: Patterns and Trends in Environmental Data
author: Lucia Scheele and Laura Vetter
format:
  html:
    code-fold: true # makes the code in the output collapsable
execute:
  warning: false # hides warnings from the generated output
  message: false        # hides messages from the generated output
lang: en  # switch to "de" if you write your report in german
bibliography: bibliography.bib
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| code-summary: preprocessing #name des Code Chunks

#loading all the data from the preprocessing files
load("my_environment.rdata")

library(tmap)
library(sf)
library(maptiles)

tmap_mode("plot") # bc mode "view" is too work intensive
#define background map
bbox <- st_bbox(selected_tracks_na_omit)
# Fetch OSM tiles
osm_bg <- get_tiles(bbox, provider = "OpenStreetMap", zoom = 16)
osm_bg_zoom <- get_tiles(bbox, provider = "OpenStreetMap", zoom = 18)

```

## Abstract

This project aims to analyze travel behaviour and clustering of travel
segments by different clustering methods and visual analysis. Parameters
like speed, acceleration, elevation and sinuosity and derived parameters
were used for clustering with kmeans and hclust. Additionally, a
segmentation by threshold (using the parameter speed) and a manual
segmentation using GIS were carried out to compare them to the
clustering. Verification took place using Chi square test. It was found
that ¨

the amount of clusters - 5 clusters, which corresponds to the actual
amount of travel modes (+standing) - hclust predicts better than
kmeans - the ward method in hclust is better than single and - xxx
showed a better (or equally good) result than manual classification. -
Verification method - ... BEARBEITEN

TO CONSIDER: Interactive maps are only reasonable for small amounts of
data (e.g. a few locations of a single wild boar). If you want to show a
large amount of data, consider making a static map instead (using
tmap_mode("plot")).

## Introduction

In order to understand travel behaviour, it is critical to analyze
movement data and be able to classify different types of movement
correctly (@sadeghian2022stepwise). However, the exact detection of the
corresponding travel modes is complex and can be time consuming.\
Generic methodologies to detect travel modes have been described using
unsupervised learning algorithms, GIS multi-criteria process (eg. by
@sadeghian2022stepwise), and supervised learning algorithms or decision
trees and confusion matrix(@shamoun2012sensor). This project aims to
answer the reasearch question of which travel modes can be
differentiated in trajectories with different travel modes within the
trajectories using hierarchical and non-hierarical clustering?
Therefore, the project focuses on a comparison of these two clustering
methods, and a comparison with a manual segmentation using ArcGIS and a
threshold segmentation using the parameter speed.

## Material and Methods

### Data

The data sample consists of movement data collected with the Tracking
App Strava, by two individuals over a timespan of April - May 2024.
Overall, 31 trips of different travel modes were recorded with the
movement types walking, running, cycling, bus, tram, train and car. For
simplification of the design and to be able to focus on the application
of the clustering methods, a sample of 2 trips with different movement
types within the trajectory were selected to be analyzed. The tracks
were selected so that all movement types standing, walking, running,
biking and tram were included. Data points were recorded every second,
resulting in a relatively high resolution of data. However, data
frequency is not consistent in all cases due to lack of signal (e.g. in
a train). At first, including contextual data such as traffic
infrastructure data by the city of Zurich (
[https://www.stadt-zuerich.ch/geodaten/](#0){.uri}) was considered. Out
of two reasons, it was decided not to include this data: 1) categorical
data is not suitable for the evaluation with the k-means clustering
algorithm and 2) the infrastructure data dit not overlap with the
trajectories in many cases and was therefore not especially meaningful.
Therefore, the focus was set on the analysis of the numerical data.

### Calulation of Parameters

The analysis was made with data of the individual fixes. instead of
using segments for the analysis of e.g. acceleration, the data of moving
windows ( X steps before and after the fix) was assigned to the
individual fix. This was done to have more information for each fix
(data point).

5 different travel modes could be identified in the samples (standing,
walking, running, tram, biking), therefore the optimal number of
clusters expected is 5, which is also the number of classes assigned in
GIS and with the speed parameter. Also, the trajectory data will be
analysed with partitioning methods to determine the number of clusters
needed, and results will be compared. Space and time of trajectories
were treated to be relative and are filtered out for the analysis. The
parameters for clustering will be calculated from the given geographical
data and include:

<ul>

<li>distance point1 - point2</li>

<li>speed p1 - p2</li>

<li>acceleration p1 - p2</li>

<li>mean speed in moving window</li>

<li>max speed in moving window</li>

<li>mean acceleration in moving window</li>

<li>max acceleration in moving window</li>

<li>elevation change in moving window</li>

<li>sinuosity in moving window</li>

</ul>

```{r }
#| eval: false
    distance = distance_by_element(geometry, lead(geometry, 1)), 
    time_diff = as.numeric(timestamp - lag(timestamp)),
    speed = distance / time_diff,
    speed_kmh = speed * 3.6, 
    acceleration = (speed - lag(speed)) / time_diff,
    avg_speed_10s = slide_dbl(speed_kmh, mean, .before = 5, .after = 5, .complete = TRUE),
    avg_speed_60s = slide_dbl(speed_kmh, mean, .before = 30, .after = 30, .complete = TRUE), 
    max_speed_10s = slide_dbl(speed_kmh, max, .before = 5, .after = 5, .complete = TRUE),
    avg_acc_10s = slide_dbl(acceleration, mean, .before = 5, .after = 5, .complete = TRUE) ,
    avg_acc_60s = slide_dbl(acceleration, mean, .before = 30, .after = 30, .complete = TRUE),
    max_acc_10s = slide_dbl(acceleration, max, .before = 5, .after = 5, .complete = TRUE),
    el_change = (elevation -lag(elevation,  10)),
    d_direct10 = distance_by_element(lag(geometry,4), lead(geometry,5)),
    d_sinu10 = rollsum(distance, 10,align = "center", fill = NA), 
    d_direct10 = case_when(d_direct10 == 0 ~ d_sinu10,
                           TRUE ~ d_direct10),
    sinuosity = d_sinu10/d_direct10   
```

The calculation of sinuosity, some adaptions needed to be made as some
values had the value "Inf". This occurs when the direct distance = 0,
meaning that start and end point of the moving window are the same. The
actual moving distance is \>1, as within the moving window movement
occured.\

actual moving distance /direct distance = x/0=Inf

There are several ways to deal with this problem:

<ul>

<li>Set inf values to the maximum value of the sinuosity</li>

<li>Filter out values completely from the table</li>

<li>Insert the actual distance traveled per 10 points,
actual/direct=x/x=1</li>

</ul>

The third method was chosen, resulting in the lowest sinuosity value
possible (=1).

### Speed Parameter Classification

In order to verify how well speed alone can already predict the
different movement types, a classification was made based on speed as
the only criterion. The different travel modes were assigned to
different speeds where:

```{r}
#speed_cluster = case_when(
   # speed_kmh == 0 ~ "1", #standing
   # speed_kmh < 5 ~ "2", #walking
   # speed_kmh >= 5 & speed_kmh < 18 ~ "3",  #running
   # speed_kmh >= 18 & speed_kmh < 30 ~ "4", # biking 
   # speed_kmh > 30 ~ "5" # tram 
   # )
```

Possible unhomogenity of some fixes could be a problem, e.g. when the
threshold is set at a point where the speed varies constantly (e.g.
speed\>5 = running, speed\<5 = walking). This issue will be resolved
with manual classification in GIS.

### Manual Classification using GIS

In order to test the results of our cluster analysis, it was aimed to
assess in how far the cluster categories are consistent with the actual
movement types conducted during the trajectory recordings. This required
two independent enrichments of the fix point data. In one step, the
information of the actual movement type had to be added to the single
trajectory fixes by using the recorder's knowledge of the track. This
step was conducted within ArcGis (Tools: Select and Calculate Field). To
help the recognition of movement type change, the fix points were
colored using the speed (km/h) variable.\
In another step, the individual cluster groups were inspected visually
and statistically to find out which movement they are likely to
represent.

### Clustering with kmeans

K means is one of the most commonly used partition based algorithms
(@yuan2017review), where k defines the number of clusters and is
identified before processing. Each partition is one cluster and contains
at least one data point, where each data point is assigned to only one
cluster. First, partitioning (determining the appropriate number of k)
is carried out using the elbow method from the package factoextra
(@ElbowMethod) and the cascade method from the package vegan.\
All parameters included in the clustering analysis are numerical:

```{r}
head(km_no_geom, 5) 
```

### Clustering with hclust

When using k-means, the iterative process results in a cluster partition
that differs with the number of requested cluster groups. In contrast,
hierarchical cluster algorithms also allow to explore how the various
clusters are related to each other with different clustering levels. The
agglomerative hierarchical clustering does this, by first assigning each
object its own cluster and iteratively joining the two most similar
clusters, until there is just one single cluster. The distance between
clusters is calculated according to a distance measure of choice
(@rdocumentationHclustFunction). Here, three typical methods for this
type of hierarchical clustering are explored: *Ward's Minimum Variance
Method*, the *complete linkage method* and *single linkage* method.\
Here, the agglomerative clustering methods were applied using *hclust*
function from the *stats* package. The required input data has to be
supplied as a dissimilarity structure. The latter was calculated from
the scaled parameter values of each observation using the *dist*
function (i.e. chord-distance). This prevents, that ensures, that
parameters with different value scales have a equal influence on the
results.

## Results

Two trajectories were chosen which include several travel modes. The
first one (light blue) is a mixture of walking, running, tram and
standing (at the tram stop). The second one (green) consists of biking,
walking and standing.

```{r}
#| fig-cap: "Map with the two Trajectories to be analyzed"

trackID_map <- tm_shape(osm_bg) +
  tm_rgb(alpha = 0.4)+
  tm_shape(selected_tracks_na_omit) + 
  tm_dots(col = "trajID", palette = "Paired", alpha = 1)

trackID_map

```

### Speed Classification

The parameters distance, speed, acceleration, elevation and sinuosity
were calculated as mentioned in the methods. The visual evaluation
showed that speed might be the most appropriate as a reference predictor
for classifying the movement types. Therefore, it was taken and a
threshold analysis was made.

```{r}
#| fig-cap: "Map 1: Speed in km/h , Map 2: Classification by Speed"


speed_map <-  tm_shape(osm_bg) +
  tm_rgb(alpha = 0.4)+
  tm_shape(selected_tracks_na_omit)+
  tm_dots(col = "speed_kmh", palette = "-RdYlGn") 


cluster_speed_map <- tm_shape(osm_bg) +
  tm_rgb(alpha = 0.4)+
  tm_shape(selected_tracks_na_omit)+ 
  tm_dots(col = "speed_cluster_name", palette = "Paired")


tmap_arrange(speed_map, cluster_speed_map )
```

The first map shows the actual speed per datafix, while the second map
shows the classification via the speed thresholds. Here, most of the
biking part and the tram are classified correctly, however the
differentiation between walking and running is not perfect and also
incorrect when the tram slows down.

### GIS Classification

To be able to differentiate between the different groups, a manual
assignation of the groups was made in ArcGIS. It works as the "reference
classification" or ground truth.

```{r}
#| fig-cap: "Map of manually assigned groups in ArcGIS"



```

### kmeans Cluster

#### Partitioning

In order to define the appropriate number of clusters, the elbow
statistics is used. The bend (like a bent elbow) in the line indicates
the number of appropriate k.

```{r}
plot_k_elbow
```

This method indicates that the best amount of clusters seems to be 2, as
the curve starts to bend there. The method was tried with several
configurations (different n-starts, leaving out parameters like
elevation or sinuosity, taking one track only), but the indicated number
for k stayed the same. As this does not correspond to the actual amount
of movement types, a different method, the cascade, was tried out. Here,
the suitability was tested for 2-8 cluster groups.

```{r}
cascade_results
```

The SSE (sum of squared errors) is a measure of how well the data points
are summarized within the clusters. The lower, the better the
clustering. In the ssi (silhouette index) the higher the value (from -1
to 1) the better the clustering. The ssi is highest for 5 groups, while
the SSE decreases constantly with the rising number of groups. This
indicates that a well chosen amount of k is 5.

#### Applying kmeans

The importance of the nstart criterium in the kmeans analysis is not to
underestimate. It is recommended to run k-means multiple times with
different random n-starts and choosing the best result, as this helps
avoid getting stuck in a poor local optimum. In the following graphic,
we can see the different results when using no predetermined n, n=20 and
n=100.

```{r}
#library("gridExtra")
#grid.arrange(plot_cluster_5, plot_cluster_5_20, plot_cluster_5_100, ncol = 3, widths = 2)
plot_cluster_5
plot_cluster_5_20
plot_cluster_5_100
```

The classification of the sample points in the point cloud, gives
insight into how many data points belong to which cluster. From this, it
is not yet possible to determine which clustering method might
correspond best to the original movement type and which amount of n
suits our classification best.

```{r}
cluster_k5_map <- tm_shape(osm_bg) +
  tm_rgb(alpha = 0.4)+
  tm_shape(selected_tracks_na_omit)+ 
  tm_dots(col = "kmeans5", palette = "Paired")

cluster_k5_20_map <- tm_shape(osm_bg) +
  tm_rgb(alpha = 0.4)+
  tm_shape(selected_tracks_na_omit)+ 
  tm_dots(col = "kmeans5_20", palette = "Paired")

cluster_k5_100_map <- tm_shape(osm_bg) +
  tm_rgb(alpha = 0.4)+
  tm_shape(selected_tracks_na_omit)+ 
  tm_dots(col = "kmeans5_100", palette = "Paired")

tmap_arrange(cluster_k5_map, cluster_k5_20_map, cluster_k5_100_map )
```

Looking at the clusters with k=5, k=5&n=20 and k=5&n=100, we can see
that the distribution of clusters is very similar, meaning that (sadly,
this is much better visible in the interactive map, but it is too heavy
to load in the html document) This means, n does not make a great
difference in the assignation of the fixes to the clusters, different to
what the plots before indicated. However, the problem of clustering
difficulties for running/walking and differentiation of tram & bike are
encountered again.

```{r}
cluster_k4_map <- tm_shape(osm_bg) +
  tm_rgb(alpha = 0.4)+
  tm_shape(selected_tracks_na_omit)+ 
  tm_dots(col = "kmeans4", palette = "Paired")
cluster_k4_map
```

Changing to k=4, running, walking and a third category can be relatively
clearly distinguished, but there is neither a distinction between tram
and bike visible. Also curves with the bike (which imply slowing down)
or tram stops are not classified correctly. For further analysis we will
therefore leave out the k=4.

### hclust Cluster

###  Hierarchical clustering (hclust)

The result of the h-clustering process is a tree of clusters which shows how the different clusters are related to eachother. 

```{r, dendogram}
par(mfrow = c(1, 3))
plot(hc_single, main = "Single", labels = NULL, sub = NULL)
plot(hc_complete, , main = "Complete")
plot(hc_ward, main = "Ward")

par(mfrow = c(1,1))
```

Depending on the input data and the question asked, the cluster can be split at different cluster levels. In our case, the cluster level of 5 is a sensible choice when considering the amount of different movement types between which we would like to distinguish. Looking at the resulting summary of the fix point distribution among the five clusters, it became apparent, that the Ward method is the only method which was able to create spherical clusters of reasonable sizes. 

```{r, cluster summary}
summary_table_hclust 
```

Since in the given tracks, the movement types were easily practiced for more than a minute, we would expect there to be more than 60 fixes per movement type cluster. As the table above shows, the clusters from the single and complete method cannot account for this distribution. This is most likely due to the effect of chaining, where single observations are considered different from all other observations. This can also be observed in the given dendrograms.  In the following, we will thus only consider the results from the ward cluster. 

```{r}
#P_ward_4

#tmap_arrange(cluster_h5_ward,cluster_k5_map )

```
Works well with differentiation between running and walking, but bike and tram look the same. 


meaning that e.g. cluster bike is closer to running and tram is closer
to train

```{r}

cluster_h5_ward <-  tm_shape(osm_bg) +
  tm_rgb(alpha = 0.4)+
  tm_shape(selected_tracks_na_omit)+ 
  tm_dots(col = "c_ward_5", palette = "RdYlGn")

tmap_arrange(cluster_h5_ward,cluster_k5_map )

```

Works well with differentiation between running and walking, but bike
and tram look the same.

### Verification with chi square test

Clusterings with different k values are not directly comparable, as they
have fundamentally different structures. Therefore, we will focus on the
comparison of k=5 only. The n criteria in kmeans does not play such a
big role in cluster assignation after all, therefore, only k=5&n=100 is
taken into account for verification. The best method to analyse several
categorical variables (cluster groups) with no supposed causality
between them is the chi square or fishers test. However, fishers test is
for smaller data samples (sample size less than 1000), therefore the chi
square is the appropriate test (Source:
https://stats.libretexts.org/Bookshelves/Applied_Statistics/Biological_Statistics\_(McDonald)/02%3A_Tests_for_Nominal_Variables/2.07%3A_Fisher's_Exact_Test
)

We will compare:

<ul>

<li>GIS - Speed</li>

<li>GIS - kmeans_100</li>

<li>GIS - hclust_ward</li>

<li>speed - kmeans_100</li>

<li>speed - hclust_ward</li>

<li>hclust_ward - kmeans_100</li>

</ul>

*X-squared*: This is the value of the chi-squared statistic. The larger
χ2, the more likely that the variables are related.

*df*: This is the number of degrees of freedom. It is typically
calculated as (number of rows-1)×(number of columns-1) in the
contingency table.

*p-value*: This is the p-value of the test. A very small p-value (in
this case extremely small) indicates that the probability of obtaining
the observed data when the null hypothesis is true is very low.

```{r}
results_df
```

The X-squared value is highest comparing the hclust and kmeans values,
but this is not of relevance. Taking GIS as the reference, from all
classifications, the hclust cluster has the highest X-squared value and
thus seems to be closest to the GIS classification. Interestingly,
kmeans is closer to the speed classification.

Lets check these results visually:

```{r}
tmap_arrange(cluster_k5names_map, cluster_h5_names_map, cluster_GIS_map)

```

```{r}
#| eval: false


cluster_GIS_map <- tm_shape(osm_bg) +
  tm_rgb(alpha = 0.4)+
  tm_shape(selected_tracks_na_omit)+ 
  tm_dots(col = "GIS_name", palette = "Paired")

cluster_k5names_map <- tm_shape(osm_bg) +
  tm_rgb(alpha = 0.4)+
  tm_shape(selected_tracks_na_omit)+ 
  tm_dots(col = "k5_names", palette = "Paired")

cluster_h5_names_map <-tm_shape(osm_bg) +
  tm_rgb(alpha = 0.4)+
  tm_shape(selected_tracks_na_omit)+ 
  tm_dots(col = "hward_names", palette = "Paired")

tmap_arrange(cluster_k5names_map, cluster_h5_names_map, cluster_GIS_map)
  
cluster_k5names_map
cluster_h5_names_map
cluster_GIS_map

```

-   Category Standing was not Identified with either of the clustering
    methods.
-   the "undefined" category is the same in both clustering methods, and
    seems to consist of points with high sinuosity or points where data
    nearby is missing (NA). No time for further investigation.
-   Differentiation between bike&tram was not possible.

## Discussion

chaining

The Chi Square results suggest that the hclust method is more appropriate to predict transport modes than the kmeans clustering. However, when analyzing visually, both methods do not take into account important
factors such as the differentiation between tram and biking. The parameter sinuosity, which was initially thought to help distinguish between the two classes, did not seem to have an effect. 


. It was observed that clustering with single tracks
works better than clustering with all tracks. 
Therefore, the applicability of this method is questionable/not suitable for bigger
datasets or datasets with different movement patterns. Furthermore, it
needs to be mentioned that for the validation with GIS, the speed
information was used as a control. Therefore, independent control of
cluster quality cannot be guaranteed.

- stationäre punkte mitgeclustert
 - spatial proximity of the data: This might have led to a disadvantage for clustering? 


### Further Research

-   At first, the turning angle was considered as an alternative
    parameter to mitigate the inaccuracies that occured when the bike
    slowed down in curves, however there was no time left to include
    this in the analysis and will therefore be left for another project.
-   Another interesting investigation would have been the possibility to
    extract the parameters by which the clusters were formed (how big of
    a weight had, e.g. the speed in comparison to the average
    acceleration). If this were possible, the weights could be adjusted
    to balance certain inaccuracies as well.
-   Further clustering, such as kmodes, which also allows
    ordinal/categorical variables to be considered, would have been of
    interest, as infrastructural data could have been included.



## Appendix

### Wordcount

<!-- after installing the wordcountaddin, remove the line "#| eval: false" -->

```{r}
#for word count:
#install.packages("pacman") 
library("pacman")
p_install_gh("benmarwick/wordcountaddin")
wordcountaddin::word_count("index.qmd") 

#knitr::spin("preprocessing.R") --> macht "run" von preprocessing file und produziert html. 

```
