---
title: "Strava_File"
format: html
---

# STRAVA PROJECT 

```{r}
install.packages("XML")
#install.packages("gitcreds")
library("readr")
library(sf)
library(ggplot2)
library(dplyr)
library("gitcreds")
library(XML) #to read the XML data of gpx files
library(leaflet) #to show in a map

laura_act <- read.csv("data/Laura_Strava/Laura_act.csv")

#laura_act <- st_as_sf(laura_act, coords = c("E", "N"), crs = 2056)
#nochmal nachschauen, wsl gar nicht nyötig

```

## Understanding GPX 
https://www.appsilon.com/post/r-gpx-files
GPX is a very common map data format used to store GPS data of routes. GPX is basically XML, therefore, we need to install the XML package. 
I tracked 7 routes of mine in total, taking different forms of movement. I want to find out, which means of transport/movement I used. First, we will load the gpx files into the project: 


```{r}
gpx_1<- htmlTreeParse(file = "data/Laura_Strava/activities/11091356418.gpx", useInternalNodes = TRUE)
gpx_2<- htmlTreeParse(file = "data/Laura_Strava/activities/11103101530.gpx", useInternalNodes = TRUE)
gpx_3<- htmlTreeParse(file = "data/Laura_Strava/activities/11116616348.gpx", useInternalNodes = TRUE)
gpx_4<- htmlTreeParse(file = "data/Laura_Strava/activities/11188517987.gpx", useInternalNodes = TRUE)
gpx_5<- htmlTreeParse(file = "data/Laura_Strava/activities/11203431760.gpx", useInternalNodes = TRUE)
gpx_6<- htmlTreeParse(file = "data/Laura_Strava/activities/11209427592.gpx", useInternalNodes = TRUE)
gpx_7<- htmlTreeParse(file = "data/Laura_Strava/activities/11239313364.gpx", useInternalNodes = TRUE)

gpx_files <- c("data/Laura_Strava/activities/11091356418.gpx","data/Laura_Strava/activities/11103101530.gpx", "data/Laura_Strava/activities/11116616348.gpx", "data/Laura_Strava/activities/11188517987.gpx", "data/Laura_Strava/activities/11203431760.gpx", "data/Laura_Strava/activities/11209427592.gpx", "data/Laura_Strava/activities/11239313364.gpx")


# I am sure there is a code to run this smoothely, but I have not encountered it yet, nor the time to invest. 
```


```{r}
#the coord info is in gpx file for every activity. 
coords <- xpathSApply(doc = gpx_1, path = "//trkpt", fun = xmlAttrs) # question: if we use the same from gpx1 for all, is that a problem?
elevation <- xpathSApply(doc = gpx_1, path = "//trkpt/ele", fun = xmlValue) # same question

df <- data.frame(
  lat = as.numeric(coords["lat", ]),
  lon = as.numeric(coords["lon", ]),
  elevation = as.numeric(elevation)
)

head(df, 10)
tail(df, 10)

plot(x = df$lon, y = df$lat, type = "l", col = "black", lwd = 3,
     xlab = "Longitude", ylab = "Latitude")
```
###CHAT GTP idea: 
make a df that has all the lat lon and elevation infos of all the trips, and give them a column with a unique ID per trip. 
then this huge df can be set into one map, structured by ID. 
therefore, I would need to add a column that has the ID 
```{r}
#CHAT GTP proposes this: 
gpx_files <- list.files("data/Laura_Strava/activities", pattern = "\\.gpx$", full.names = TRUE)

head(gpx_files)

process_gpx <- function(file, id) {
  doc <- xmlTreeParse(file, useInternalNodes = TRUE)
  coords <- xpathApply(doc, "//trkpt", function(x) as.numeric(xmlAttrs(x)[c("lat", "lon")]))
  elevation <- xpathSApply(doc, "//trkpt/ele", xmlValue)
  df <- data.frame(
    lat = coords[, 1],
    lon = coords[, 2],
    elevation = as.numeric(elevation),
    object_id = id
  )
  return(df)
}

all_data <- do.call(rbind, lapply(seq_along(gpx_files), function(i) {
  process_gpx(gpx_files[i], i)
}))

head(all_data)

```



## loading several files into one Map
https://stackoverflow.com/questions/54726758/merging-multiple-gpx-files-into-a-single-gpx-file-with-multiple-tracks

```{r}
gpx_files <- c("data/Laura_Strava/activities/11091356418.gpx","data/Laura_Strava/activities/11103101530.gpx", "data/Laura_Strava/activities/11116616348.gpx", "data/Laura_Strava/activities/11188517987.gpx", "data/Laura_Strava/activities/11203431760.gpx", "data/Laura_Strava/activities/11209427592.gpx", "data/Laura_Strava/activities/11239313364.gpx")

#install.packages("plotKML") library("plotKML") -- funktioniert nicht auf neuem R 
install.packages("sf")
library(sf)

# Create empty list for storing .gpx files.
list_gpx <- list()

# Loop to read files, store in a list with name:
for (i in seq_along(gpx_files)){
  list_gpx[[i]] <- readGPX(gpx_files[1])
  names(list_gpx)[[i]] <- gpx_files[i]
}

# Loop through list_gpx, get track df, clean up columns, and save in list. 
# Empty list for tracks. 
track_list <- list()

# Loop
for (i in 1:length(list_gpx)){
  track_list[[i]] <- do.call(cbind,list_gpx[[i]]$tracks[[1]])[,c(1:4)]
  if (grepl("Run",colnames(track_list[[i]]))==TRUE){
    track_list[[i]]$activity <- rep("Run",nrow(track_list[[i]]))
  }else{
    track_list[[i]]$activity <- rep("Bike",nrow(track_list[[i]]))
  }
  names(track_list[[i]]) <- c("lon","lat","ele","time","activity")
}

# Merge dataframes in track_list.
data <- do.call(rbind,track_list)


```

```{r}
```




## colouring etc
```{r}
library(leaflet)

leaflet() %>%
  addTiles() %>%
  addPolylines(data = df, lat = ~lat, lng = ~lon, color = "#000000", opacity = 0.8, weight = 3)


get_color <- function(elevation) {
  if (elevation < 500) {
    return("green")
  }
  if (elevation < 1000) {
    return("yellow")
  }
  if (elevation < 1500) {
    return("orange")
  }
  return("red")
}




# New dataset with the new variable for color
df_color <- df %>%
  rowwise() %>%
  mutate(color = get_color(elevation))

df_color$last_color <- dplyr::lag(df_color$color)

# Map
map <- leaflet() %>% addTiles()
for (color in levels(as.factor(df_color$color))) {
  map <- addPolylines(map, lat = ~lat, lng = ~lon, data = df_color[df_color$color == color | df_color$last_color == color, ], color = ~color)
}
map
```
